Testing and Isolation Efficacy:\\Insights from a Simple Epidemic Model

### 2021-07-14 SIR_manuscript.tex
a few more tweaks

However, we did find two surprising patterns: longer delays in returning tests can \emph{reduce} epidemic spread, and increasing testing rates can \emph{increase} spread.

@@

The other counterintuitive result from our analysis is that, for sufficiently high testing intensity $\rho$, further increasing testing intensity can actually \emph{increase} $\Rnum$ (e.g., \fref{pan2}(b), upper right panel [$\theta\_c=1$, $\theta\_w=0$]). This phenomenon can occur because we are considering the DFE in the presence of testing; thus there is an equilibrium distribution of susceptibles between the $S\_n$ (untested) and $S\_u$ (waiting) compartments even as the disease approaches extinction. A higher rate of testing leads to a greater proportion of individuals waiting for negative tests at the DFE. If infected, individuals in this group will take longer to be tested again and to subsequently isolate (because they must wait for their negative tests to be returned before being tested again). If isolation in this group ($\theta\_w$) is low, this effect can under some (relatively rare) circumstances (high $\theta\_c$, low $\omega$, high $\rho$) allow $\Rnum$ to increase with testing intensity. 
We can show that this phenomenon occurs \emph{only} under targeted testing ($w_I > w_S$), but we have not yet found a simple explanation of why it cannot occur without it. This phenomenon is also unlikely to occur in the real world. In particular, it depends on levels of testing that are unrealistically high (at least in large, general-population settings).

@@

Although testing and tracing is a key part of infection control strategies, mathematical epidemiologists have typically analyzed it with detailed models designed to inform particular public health efforts \new{\citep{endo2020implication,hellewell2020feasibility,jenness2020modeling}}, rather than analyzing simple but general models of the feedback between testing and transmission dynamics.
\new{There have been several modeling studies of testing and tracing dynamics and their interaction with epidemiological dynamics. In the context of repeated screening and random testing of isolated populations (such as a university), \cite{bergstrom2020frequency} provided some analytical results which suggest that proactively screening asymptomatic individuals and isolating confirmed-positive cases would reduce the spread of the disease. 
\cite{rogers2021high} simulated a SEIR model with testing and isolation, and suggest a strategy of rapid testing with antigen tests and the subsequent isolation of confirmed-positive individuals.
H \cite{friston2021testing} model the effects of self-isolation on testing and tracing with a focus on projections under different testing and tracing scenarios. They conclude that the emergence of a second wave depends almost primarily on the rate at which immunity is lost and that it is necessary to track asymptomatic individuals in order to control the outbreak. 
Our modeling approach differs from earlier work in that it examines the effects of test-return rates and of different levels of testing focus, from random to highly targeted. We hope this paper will inspire further explorations of the fundamental properties of epidemic models that incorporate explicit testing processes.}

@@

It remains hard to show that the reproduction number $\Rnum$ is decreasing with respect to \percap testing intensity, $\rho$, and the speed of the test return, $\omega$, for the feasible ranges of the parameters, that is
\begin{align}
\label{cond}
& \omega>0, \\
& 0 \leq \rho<\omega,\\ 
& 0 \leq \theta\_w \leq \theta\_c \leq 1, \\
& \testtarget\geq 1.
\end{align}
In realistic cases the testing rate $\rho$ is very small (i.e., only a small fraction of the population can be tested every day); it is thus reasonable to use a linear approximation of $\Rnum$ for $\rho \ll 1$ to analyze the behaviour of $\Rnum$ with respect to $\omega$ (see section \appref{omega}). 
In the next section we provide an equivalent representation of $\Rnum$ in order to show that increasing testing intensity typically decreases $\Rnum$.

@@

An example of this countervailing effect of $\phi$, and consequently $\rho$, on $\Rnum$ occurs when $\theta\_w=0$ and $\theta\_c=1$.
This is illustrated in the top-right panel of the \fref{pan2} panel (b), where the strength of isolation for awaiting people is the least, but the most for the confirmed cases. In this case, simplifying $a_3$ in Eq.~\eqref{eq:abcd2} gives
\begin{equation}
  \begin{split}
    a_3 & =\testtarget \omega \gamma p_I\left((\omega+\gamma)-\testtarget(\omega p_I+\gamma)\right)  \\
    & \propto  \omega \left( 1- \testtarget p_I \right) + \gamma \left( 1- \testtarget \right)  .
  \end{split}
\end{equation}
If $p_I>0$, then $a_3<0$ for sufficiently targeted testing (i.e. when $\testtarget p_I >1$; because
$p_I \leq 1$, \testtarget is always $\geq \testtarget p_I$).
%% this LOOKS like we can get a3<0 even if pI=0 for targeting > 1 and sufficiently large gamma/omega,
%% but that seems crazy so I'm probably missing something
When the test is perfectly sensitive ($p_I=1$), $a_3<0$ as long as $\testtarget>1$.
Under either of these conditions, there exists a range for $\rho$ over which $\pder\Rnum{\rho}\leq 0$.  
Note that $\rho$ and $\omega$ have a similar mechanism in delaying people to get into $I\_c$, thus we would expect to see the non-trivial countervailing effect of these two parameters on $\Rnum$.

@@

Having presented the formal analysis, we now concern ourselves with its biological interpretations. We begin by interpreting \ref{case_1_ineq}, under which returning tests more rapidly is \emph{always} harmful. Notice that the ratio $\frac{\theta\_c}{\theta\_w}$ is simply a measure of how much more strongly individuals self-isolate when they test positive compared to when waiting for tests. Since the rate of test return directly influences the rate at which individuals change from a waiting state to a confirmed-positive state, it is intuitive that $\frac{\theta\_c}{\theta\_w}$ would appear in \ref{case_1_ineq}. Next, note that the left-hand side increases when test sensitivity decreases and when targeting of positive individuals is poor. This is consistent with our intuition: a false negative that is returned more rapidly may encourage an infectious individual to not carefully self-isolate, thus increasing transmission. Likewise, if individuals tested are mainly susceptible (rather than infectious), then returning tests more slowly would encourage them to self-isolate for longer while awaiting test results. Having understood the role of each of the parameters in \ref{case_1_ineq}, a holistic interpretation of this inequality is that returning tests more slowly is helpful when the benefit of extended self-isolation (due to awaiting test results) outweighs the benefit of identifying positive cases.

>>>>

However, we did find two surprising patterns: under some conditions longer delays in returning tests can reduce epidemic spread, and increasing testing rates can increase spread.

@@

The other counterintuitive result from our analysis is that, for sufficiently high testing intensity $\rho$, further increasing testing intensity can actually \emph{increase} $\Rnum$ (e.g., \fref{pan2}(b), upper right panel [$\theta\_c=1$, $\theta\_w=0$]). This phenomenon can occur because we are considering the DFE in the presence of testing; thus there is an equilibrium distribution of susceptibles between the $S\_n$ (untested) and $S\_u$ (waiting) compartments even as the disease approaches extinction. A higher rate of testing leads to a greater proportion of individuals waiting for negative tests at the DFE. If infected, individuals in this group will take longer to be tested again and to subsequently isolate (because they must wait for their negative tests to be returned before being tested again). If isolation in this group ($\theta\_w$) is low, this effect can under some (relatively rare) circumstances (high $\theta\_c$, low $\omega$, high $\rho$) allow $\Rnum$ to increase with testing intensity. 
We can show that this phenomenon occurs \emph{only} under targeted testing ($w_I > w_S$), but we have not yet found a simple explanation of why it cannot occur under random (unfocused) testing. This phenomenon is also unlikely to occur in the real world. In particular, it depends on levels of testing that are unrealistically high (at least in large, general-population settings).

@@

Although testing and tracing is a key part of infection control strategies, mathematical epidemiologists have typically analyzed it with detailed models designed to inform particular public health efforts \new{\citep{endo2020implication,hellewell2020feasibility,jenness2020modeling}}, rather than analyzing simple but general models of the feedback between testing and transmission dynamics.
\new{There have been several modeling studies of testing and tracing dynamics and their interaction with epidemiological dynamics. In the context of repeated screening and random testing of isolated populations (such as the members of a university), \cite{bergstrom2020frequency} provided analytical results quantifying the effects that proactive screening of asymptomatic individuals and isolation of confirmed-positive cases could have in reducing the spread of disease. 
\cite{rogers2021high} simulated a SEIR model with testing and isolation; they similarly suggest a strategy of rapid testing with antigen tests and the subsequent isolation of confirmed-positive individuals.
\cite{friston2021testing} model the effects of self-isolation on testing and tracing with a focus on projections under different testing and tracing scenarios. They conclude that the emergence of a second wave depends almost primarily on the rate at which immunity is lost and that it is necessary to track asymptomatic individuals in order to control the outbreak. 
Our modeling approach differs from these previous efforts in that it examines the effects of test-return rates and of different levels of testing focus, from random to highly targeted. We hope this paper will inspire further explorations of the fundamental properties of epidemic models that incorporate explicit testing processes.}

@@

It remains difficult to show that the reproduction number $\Rnum$ is decreasing with respect to \percap testing intensity, $\rho$, and the speed of the test return, $\omega$, for the feasible ranges of the parameters, that is
\begin{align}
\label{cond}
& \omega>0, \\
& 0 \leq \rho<\omega,\\ 
& 0 \leq \theta\_w \leq \theta\_c \leq 1, \\
& \testtarget\geq 1.
\end{align}
In realistic cases the testing rate $\rho$ is very small (i.e., only a small fraction of the population can be tested every day); it is thus reasonable to use a linear approximation of $\Rnum$ for $\rho \ll 1$ to analyze the behaviour of $\Rnum$ with respect to $\omega$ (see section \appref{omega}). 
In the next section we provide an equivalent representation of $\Rnum$ in order to show that increasing testing intensity typically decreases $\Rnum$.

@@

An example of this countervailing effect of $\phi$, and consequently $\rho$, on $\Rnum$ occurs when $\theta\_w=0$ and $\theta\_c=1$.
This is illustrated in the top-right panel of the \fref{pan2} panel (b), where the strength of isolation for awaiting people is the least, but the most for the confirmed cases. In this case, simplifying $a_3$ in Eq.~\eqref{eq:abcd2} gives
\begin{equation}
  \begin{split}
    a_3 & =\testtarget \omega \gamma p_I\left((\omega+\gamma)-\testtarget(\omega p_I+\gamma)\right)  \\
    & \propto  \omega \left( 1- \testtarget p_I \right) + \gamma \left( 1- \testtarget \right)  .
  \end{split}
\end{equation}
If $p_I>0$, then $a_3<0$ for sufficiently targeted testing (i.e. when $\testtarget p_I >1$; because
$p_I \leq 1$, \testtarget is always $\geq \testtarget p_I$).
%% this LOOKS like we can get a3<0 even if pI=0 for targeting > 1 and sufficiently large gamma/omega,
%% but that seems crazy so I'm probably missing something
When the test is perfectly sensitive ($p_I=1$), $a_3<0$ as long as $\testtarget>1$.
Under either of these conditions, there exists a range for $\rho$ over which $\pder\Rnum{\rho}\leq 0$.  
Because increasing values of $\rho$ and $\omega$ both delay the rate at which individuals flow to the $I\_c$ compartment, it is reasonable that increasing either value could (under appropriate circumstances) increase $\Rnum$.

@@

Having presented the formal analysis, we now concern ourselves with its biological interpretations. We begin by interpreting \ref{case_1_ineq}, under which returning tests more rapidly is \emph{always} harmful. Notice that the ratio $\frac{\theta\_c}{\theta\_w}$ is simply a measure of how much more strongly individuals self-isolate when they test positive compared to when waiting for tests. Since the rate of test return directly influences the rate at which individuals change from a waiting state to a confirmed-positive state, it is intuitive that $\frac{\theta\_c}{\theta\_w}$ would appear in \ref{case_1_ineq}. Next, note that the left-hand side increases when test sensitivity decreases and when targeting of positive individuals is poor. This is consistent with our intuition: a false negative that is returned more rapidly will allow an infectious individual to relax their self-isolation, thus increasing transmission. Likewise, if individuals tested are mainly susceptible (rather than infectious), then returning tests more slowly would encourage them to self-isolate for longer while awaiting test results. Having understood the role of each of the parameters in \ref{case_1_ineq}, a holistic interpretation of this inequality is that returning tests more slowly is helpful when the benefit of extended self-isolation by infected individuals awaiting test results outweighs the benefit of identifying positive cases.
